Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-06 paper.tex
bib cleanup + reply to TS comment
Raph's comments and suggestions

@@

\subsubsection{Inter-Workgroup Barriers}
The virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier. As previously mentioned, workgroups cannot context switch. Although contemporary GPU hardware~\cite{} supports workgroup preemption for prioritizing real-time tasks or managing multi-process workloads, this functionality is deliberately excluded from the GPU programming model. This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip. In the oversubscription paradigm, where the number of workgroups is theoretically unbound and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups. Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points. However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.

@@

\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subset of elements in the sequence. If the subset includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans depart from early PRAM-like scans~\cite{} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers. Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory. Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{}. Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization. This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization. \thomas{I have written this paragraph like twenty times now, and it's still odious.}

@@

In any \emph{Chained-Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. During this waiting period, the workgroup is continually scheduled, and the \emph{progress} of other workgroups must be guaranteed to prevent the spinning workgroup from starving its dependent predecessors of execution, which could result in deadlock.

@@

@@

@@

However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and ARM GPUs do not. Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient—it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

@@

@@

@@

\subsection{Earlier Attempts at Portability}
\john{I'd like Raph to chime in here. What I worry about in the way this subsection is presented is that a reviewer will say ``Oh, the fallback technique was already invented in 2021, all this paper does is makes it a little better''. On one hand I totally want Raph to get credit for this cool idea. On the other hand, writing this subsection in this way, as a citable previous publication, significantly reduces the size of the advance that this paper we're writing hopes to capture. If I'm being selfish and trying to make this paper look as good as possible, I downplay the previous work a little and don't present it as if it was a previous peer-reviewed publication. At the least I want to use the words ``blog post'' in the text here.}
Levien and Naur~\cite{Raph2021} made the first attempt at implementing a \emph{Decoupled-Lookback} approach without FPG requirements in \emph{Scalar Fallback}. Each time a workgroup spins on a preceding tile, it advances a \emph{fallback} operation by incrementally consuming the blocking tile---one element at a time. This eliminates the \emph{mutual exclusion} condition necessary for deadlock, as workgroups can operate on data from other tiles. While this may introduce redundant work when a tile is late rather than truly blocking, it guarantees termination by allowing any workgroup to complete a blocking tile’s reduction, breaking inter-workgroup dependencies.

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the lookback thread performs a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Additionally, since only one thread loads elements, the associated cache line is severely underutilized. Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. \john{If we can, we would like to [politely] say that this previous work was missing a key insight, e.g., while fallback can address the portability problem, \ldots\ Would we say that it's not decoupled in the way that the technique we present in this paper is? Or is the only deficiency the choice of a serial fallback?} While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.

\section{Decoupled Fallback}
\emph{Decoupled Fallback} separates fallback operations from the lookback process, enhancing the \emph{Scalar-Fallback} technique with significantly improved fallback performance. As an extension of the \emph{Chained-Scan} architecture, it retains the advantages of high speed and oversubscription, aligning with our design requirements.

Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to preclude deadlocks. \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking. While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible. Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, shifting rather than eliminating deadlock risks. Removing \emph{resource holding}---allowing workgroups to discard \emph{blocked} work tiles---is theoretically possible but strictly inferior to removing \emph{mutual exclusion}, as it incurs $t$ extra memory operations (discard-read-read instead of read).
\begin{algorithm}
  \small
  \SetAlgoLined
  \KwGlobalIn{%
    $\textit{in}[]$,
    $\textit{out}[]$,
    $\textit{spine}[]$,
    $\textit{tileBump}$,
  }
  \KwSharedIn{%
    $\textit{wg\_partials}[]$,
    $\textit{wg\_fallback}[]$,
    $\textit{wg\_prev\_reduction}$,
    $\textit{wg\_broadcast}$,
    $\textit{wg\_control}$,
  }
  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
  \KwConstants{Vectors per thread $\textit{p}$}

@@

To optimize the fallback operation, we employ three strategies. Although \emph{Chained-Propagation} is performed by a single thread or subgroup, for performance reasons we want the entire workgroup to participate in the fallback. (1) First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary. (2) Next, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback. While this approach risks redundant fallbacks---since a late tile is indistinguishable from a genuinely blocking one---exposing the maximum spin count as a tunable parameter helps minimize false positives. \john{I can see it is sensible to have a maximum spin count but I do not see how the action of exposing the maximum spin count to the programmer minimizes false positives.} (3) Finally, after completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.

@@

\john{In this item + the previous item, I bet there's some ordering constraints, either for correctness or for performance. For instance, I bet the highest priority is to add the previous reduction to my reduction and posting the result, since that propagation speed across tiles is probably really important for performance. Perhaps worth pointing out. Does the flag-unset have any ordering constraints in terms of correctness?}
\thomas{Last phase may still need revision in accordance with above feedback. . . I have deliberately tried to avoid mentioning barriers to keep the section succinct.}
\john{I agree with your succinct strategy. I like that you now say reduction-propagation is the first priority, and why.}
\thomas{No furher revisions required/OK to remove comments?}.

\subsection{Tile State Implementation}
Since workgroups must observe updates from one another, but WebGPU lacks a \emph{volatile}-like qualifier, we rely on atomics to ensure visibility and consistency across workgroups. Because a tile may be in three possible states, two bits are required to record its state alongside the bits needed to store the tile's reduction value. Previous \emph{Chained-Scan} architectures relied on 64-bit atomics and/or memory fences to ensure a consistency between state and value, but WebGPU’s lack of a memory model \john{I'm not sure I would say ``lack of a memory model''. It's not the (lack of) memory model that's the problem. It's specifically that we don't have the operations that give us the behavior that we require. If we had the right operations that did the right thing but didn't have a memory model, we'd be fine. We want to specifically call out what is missing and what that means, and do so as generically as possible (not ``we are missing this specific atomic call'' but instead ``we are missing some sort of atomic call that has the following behavior''). Above, we want to not only say that we don't have a volatile qualifier but also both what that means and why it would be specifically useful if we had it. Doing the latter two things gives valuable ammunition to WebGPU and GPU architects that want to improve the API and hardware.} and 64-bit atomic support makes both infeasible. Instead, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state. \john{I would try to add a sentence or two on why we can't do the obvious thing---having the state in one word and the reduction value in a different word---and why the split value avoids that problem. This is not obvious to the reader.}
\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{graphics/split.pdf}
\end{figure}

@@

Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state. When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}. WebGPU does not guarantee the correctness of atomic \emph{Compare-and-Swap} across all devices, inhibiting its use. \john{Is CAS the traditional/common way to implement this? If so, say so, and if you can, provide a cite, even if it's not a formal cite but just the name of another package that uses this strategy.} Instead, we encode the tile state in the most significant bits of the split value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}. This allows an atomic \emph{maximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.

>>>>

\raph{I think "NVIDIA Tesla: A Unified Graphics and Computing Architecture" Lindholm 2008 might be a good choice for the above cite.}

@@

\subsubsection{Inter-Workgroup Barriers}
The virtualization of processors introduces an emergent high-level limitation: the absence of an inter-workgroup barrier. As previously mentioned, workgroups cannot context switch. Although contemporary GPU hardware~\cite{} supports workgroup preemption for prioritizing real-time tasks or managing multi-process workloads, this functionality is deliberately excluded from the GPU programming model. This stems from the prohibitively high overhead in latency and storage associated with moving workgroup contexts on and off chip. In the oversubscription paradigm, where the number of workgroups is theoretically unbounded and typically far exceeds the available multiprocessors, every invocation of such a barrier would necessitate context-switching between all launched workgroups. Due to this high cost, no GPU programming environment furnishes developers with a true unbounded-launch inter-workgroup barrier.\footnote{CUDA recently introduced \emph{Thread Block Cluster} synchronization~\cite{NvidiaCudaGuide}, but it is almost certainly designed to operate within the \emph{Persistent-Thread}~\cite{gupta2012} paradigm, rather than serving as a barrier across an unbounded launch.} Instead, when inter-workgroup synchronization is necessary, developers must rely on kernel launches as synchronization points. However, kernel launches are suboptimal as barriers: they incur overheads due to the heterogeneous nature of the operation, and all execution context is lost between launches.

\raph{I think the above is conflating memory barriers with control barriers. All modern GPU shading languages (since Metal 3.2) provide memory barriers with inter-workgroup (device) scope, but the above paragraph makes more sense interpreted as a control barrier. It might also be worth mentioning that kernel launches do not necessarily sychronize; in Vulkan it's really the pipeline barrier that does the synchronization. I can take a whack at re-writing this.}

@@

\subsection{The Scan Primitive}\label{sec:the-scan-primitive}
The study of \emph{scan} (\emph{parallel prefix}) networks traces back to the design of carry-lookahead adder circuits and beyond~\cite{5219801,10.5555/1098666}. A scan is typically defined on a monoid \( M \), characterized by a binary reduction operator \( \oplus \) and an identity element \( \varepsilon \). The binary operator \( \oplus \) satisfies the closure property \( \forall a, b \in M, \ (a \oplus b) \in M \) and has an identity element \( \exists \varepsilon \in M, \ \forall a \in M, \ \varepsilon \oplus a = a \). To allow parallelization, \( \oplus \) must be associative, but it is not necessarily commutative, as demonstrated in structures like the \emph{bicyclic semigroup}~\cite{}\footnote{Despite its name, this is a monoid.} and Rabin-Karp's \emph{second fingerprinting method for string matching}~\cite[Section 6]{Karp:1987:ERP}. In a scan, the result at the \( n \)-th element is the reduction of the preceding subsequence of elements. If the subsequence includes the \( n \)-th element, it is called \emph{inclusive}; if it excludes the \( n \)-th element, it is called \emph{exclusive}. The most common scan type is the prefix sum, where \( \oplus \) is addition. For example:
\[
  x = [x_1, x_2, x_3, \dots, x_n] \ \ \ \ \ \ \ y = [1, 1, 1, 1, 1]
\]
\[
  \text{InclusiveScan}(x, \oplus) = [x_1, x_1 \oplus x_2, x_1 \oplus x_2 \oplus x_3, \dots, x_1 \oplus x_2 \oplus \cdots \oplus x_n]
\]
\[
  \text{InclusiveScan}(y, +) = [1, 2, 3, 4, 5]
\]
Due to its significance in circuits and as a fundamental algorithmic primitive, scan has been extensively studied in both electrical engineering and computer science. Harris~\cite{1292373} offers a taxonomy that relates depth, fanout, and wire tracks, while Hinze~\cite{10.1007/978-3-540-27764-4_11} develops an algebraic framework for scans. Snir~\cite{10.1016/0196-67748690003-9} proved that depth $d$ and size $s$ are related by $s + d \ge 2n - 2$, and Fich~\cite{10.1145/800061.808738} proved that among minimum depth scans Ladner-Fischer~\cite{10.1145/322217.322232} networks have optimal size. Blelloch~\cite{Blelloch:1989:SAP} adapted scan to the PRAM model and popularized its use as an algorithmic primitive. Finally, Merrill and Garland~\cite{Merrill2016} provide a review of GPU scan implementations, complementing earlier work by Merrill and Grimshaw~\cite{Merrill2009}.

\subsection{Evolution of Inter-Workgroup Scan Architectures}
Contemporary GPU scans depart from early PRAM-like scans~\cite{Blelloch:1989:SAP} by leveraging the GPU memory hierarchy, which consists of progressively faster but increasingly private tiers. Efficient scans integrate fine-grained intra-workgroup (\emph{local}) strategies in shared memory and registers with coarse-grained inter-workgroup (\emph{global}) strategies in global memory. Given their low arithmetic intensity, local scans are inherently memory-bound, as demonstrated by Merrill and Grimshaw~\cite{Merrill2009}. Consequently, recent scan optimizations have shifted toward refining global strategies, particularly inter-workgroup coordination and synchronization. This section reviews three key approaches---\emph{Scan-then-Propagate}, \emph{Reduce-then-Scan}, and \emph{Chained-Scan}---analyzing their handling of inter-workgroup dependencies and efficiency in minimizing global memory access while maximizing bandwidth utilization. \thomas{I have written this paragraph like twenty times now, and it's still odious.} \raph{It's fine. Feel free to delete these notes in next round.}

@@

In any \emph{Chained-Scan}, once a workgroup computes its local scan, it must complete two objectives: communicate its reduction to successor tiles and resolve dependencies by incorporating the reductions of all preceding tiles. This process, which we refer to as \emph{Chained Propagation}, defines the set of inter-workgroup techniques that facilitate dependency resolution. In \emph{StreamScan}, these two objectives are combined into a single phase, where each workgroup waits for its immediate predecessor to complete, ingests the predecessor's reduction, and posts the resulting \emph{inclusive} reduction for its immediate successor to consume. This approach depends on a \emph{forward progress guarantee} of the GPU's execution model. Given such a guarantee, the predecessor will eventually produce a result and broadcast it. On devices lacking such a guarantee, the workgroup may wait indefinitely, resulting in deadlock.

@@

\raph{I'm having trouble with the above paragraph. By my read, Reduce-Then-Scan avoids inter-workgroup synchronization other than the pipeline barriers between kernel launches. Chained-Scan and Decoupled-Lookback both rely on inter-workgroup synchronization. Not sure why the "in some cases" qualifier either; seems to me you need synchronization except in the trivial case the entire array fits in one workgroup.}

@@

\raph{"Fixed launch" means launching a fixed number of workgroups and having each workgroup process multiple blocks, generally total blocks / num workgroups? I think I get the argument here but it's more subtle and not an inherent problem with RTS. You make a choice: either you do fixed launch and have to tune, *or* you can do a fixed number of blocks per workgroup, but in the latter case you need to need vary the number of dispatches based on total number of blocks. That's annoying but not a showstopper.}

@@

However, while \emph{Chained-Scan} with \emph{Decoupled Lookback} appears to be the ideal scan architecture in both theoretical and practical terms, it has a critical limitation: its reliance on a forward-progress guarantee. While NVIDIA GPUs provide this guarantee, data collected by Sorensen et al.~\cite{sorensen2021} shows that Apple’s M-series and ARM GPUs do not. Without forward progress guarantees, \emph{Chained-Scan} is not merely inefficient---it risks indefinite deadlock, causing the application to freeze and rendering the kernel unable to terminate, requiring external intervention to recover.

@@

\raph{I think the deep-dive into the nature of deadlock can be trimmed back here, but chatting with John, such editing for concision can be done later, when we have a better sense about page count.}

@@

\raph{Answering John's question here. Short answer no. One mental model that would illustrate this is a workgroup being preempted from a multiprocessor for system reasons, for example to perform compositing for scanout. I brought this up with a GPU engineer and they said this wasn't actually the reason for lack of FPG on their hardware, but wasn't able to explain the true story for proprietary reasons. In any case, while I am fine to a large extent talking operationally about scheduling etc, at the end of the day you need a formal model that tells you which computations will terminate. That's what Tyler's work is about and I think it's best to cite it and not try to re-litigate it, especially in oversimplified form. Also, taking another look at "Specifying and Testing," I notice that they use "livelock" but not deadlock to describe situations where system progress depends on FPG. I'm not finding a cite for this now, but if memory serves, a deadlock condition is one in which no progress is possible under *any* scheduling condition, while in livelock there is some scheduler condition that could resolve the livelock, but it is not resolved under all such conditions.}

@@

\subsection{Earlier Attempts at Portability}
Our first attempt \raph{we'll add the cite post-review} at implementing a \emph{Decoupled-Lookback} scan without relying on FPG was \emph{Scalar Fallback}. Each time a workgroup spins waiting on the reduction from a preceding tile, it advances a \emph{fallback} operation by processing one scalar element from that tile. This eliminates the possibility of deadlock because the number of such operations is bouded; the wait will eventually terminate, possibly at the expense of doing redundant work.

\emph{Scalar Fallback} suffers from two primary inefficiencies. First, the fallback reduction is performed serially by a single thread, progressing only when the lookback thread performs a waiting spin. This results in a $t$-sized blocking tile requiring $t$ spins to complete. Second, once the fallback completes, the lookback thread does not post its result to device memory, forcing every blocked workgroup to redundantly recompute the fallback. While \emph{Scalar Fallback} achieved speed-of-light performance on hardware with FPG and executed correctly without it, its performance on the latter was inferior to \emph{Reduce then Scan}.

\raph{I took out the statement about cache underutilization because I'm not sure it's true (L1 cache is shared by all subgroups in a multiprocessor) and I don't think it's essential to our argument. Also deleting John's comment; I don't think this is a case where we need to prove novelty over a previous publication, instead we just subsume it.}

\section{Decoupled Fallback}
\emph{Decoupled Fallback} addresses both inefficiencies. The key advance is to utilize the parallelism available in the workgroup to perform the fallback computation, rather than performing it sequentially. To do so requires nontrivial coordination within the workgroup, given that the \emph{Chained-Propagation} logic is performed by a single thread.

\raph{This paragraph feels to me like it can be cut, but I won't insist on it.} Like \emph{Scalar-Fallback}, we choose \emph{mutual exclusion} as our approach to preclude deadlocks. \emph{Dependency relationships} and \emph{no preemption} are immutable characteristics of scan and the GPU programming model, eliminating them as viable avenues for breaking deadlocking. While it may be possible to break \emph{circular ordering} by designing tile sizes such that workgroups are mapped one-to-one to multiprocessors, doing so portably is infeasible. Moreover, vendor guidance on progress guarantees among multiprocessors is also lacking, shifting rather than eliminating deadlock risks. Removing \emph{resource holding}---allowing workgroups to discard \emph{blocked} work tiles---is theoretically possible but strictly inferior to removing \emph{mutual exclusion}, as it incurs $t$ extra memory operations (discard-read-read instead of read).
\begin{algorithm}
  \small
  \SetAlgoLined
  \KwGlobalIn{%
    $\textit{in}[]$,
    $\textit{out}[]$,
    $\textit{spine}[]$,
    $\textit{tileBump}$,
  }
  \KwSharedIn{%
    $\textit{wg\_partials}[]$,
    $\textit{wg\_fallback}[]$,
    $\textit{wg\_prev\_reduction}$,
    $\textit{wg\_broadcast}$,
    $\textit{wg\_control}$,
  }
  \KwGlobalOut{$\textit{out}[]$ inclusive scan of $\textit{in}[]$}
  \KwConstants{Vectors per thread $\textit{p}$}

@@

To optimize the fallback operation, we employ three strategies. (1) First, a shared memory \emph{control flag} enables threads performing the lookback to signal the entire workgroup when a fallback is necessary. (2) Next, we limit the number of spins a thread can perform while waiting on a tile; exceeding this threshold triggers a fallback. While this approach risks redundant fallbacks---since a late tile is indistinguishable from a genuinely blocking one---exposing the maximum spin count as a tunable parameter helps minimize false positives. \john{I can see it is sensible to have a maximum spin count but I do not see how the action of exposing the maximum spin count to the programmer minimizes false positives.} (3) Finally, after completing a fallback, the workgroup attempts to post the reduction result of the blocking tile, preventing redundant fallbacks by subsequent workgroups.

\raph{To address John's comment: increasing the spin count reduces false positives; a false positive can be defined as the preceding tile posting the result late, meaning that there is some spin count for which it would proceed. However, it also delays progress in the true positive case. The deeper way to talk about this is to engage the cost model, which is done to some extent in the paragraph below. The cost of a false positive includes increased memory traffic and also compute opportunity cost of other workgroups which can be scheduled on the same multiprocessor. Indeed, with sufficient occupancy the cost of a workgroup spinning (or making scalar progress) is minimal. So optimal performance really depends on the *distribution* of spin counts. I put all this in a comment but think we can say it in the paper.}

@@

\subsection{Tile State Implementation}
In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models. \raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper". I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

Combined control and data synchronization can be accomplished by packing both the tile state and the reduced value into a single word, using relaxed atomic store to send the message, and relaxed atomic load to receive the message, spinning as necessary until the tile state is updated. The semantics of atomics guarantee, even with relaxed memory ordering, that the data is fresh. However, if the tile state and data are stored in two different memory locations, then relaxed atomics are not sufficient and some other sychronization is necessary. On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee. \raph{Here's a good place to cite the MP litmus test because that's exactly what it's about.} Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes. A device-scoped memory barrier has similar problems. While it would provide the necessary guarantee, it is also not available in WebGPU (or in versions of Metal prior to 3.2), and is expensive.

Without some other mechanism, the requirement to pack tile state and data into a single word limits the data to 30 bits (assuming 2 bits for the tile state). That, in turn, makes it impossible to represent common ordinary 32-bit floating point values without loss of precision. While some GPUs support 64 bit atomics (and it is likely that WebGPU will include them as an optional extension), that is also sadly not portable, and merely raises the payload limit to 62 bits without eliminating it altogether.

A key innovation of this paper is to \emph{split} the reduced values into multiple atomics, each carrying its own copy of the control flags. The message is received when relaxed atomic loads of all the parts agree on their control flag. This way, the payload is reconstructed intact, even in the absence of any guarantees from the memory model stronger than relaxed atomics.

In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{graphics/split.pdf}
\end{figure}

@@

Fallbacks require a tile state representation and an atomic operation that can safely transition a tile to \emph{Ready} without overwriting a more advanced state. When an update attempt is made, a blocking tile may be in any of the three states due to either of two scenarios: (1) a false-positive fallback on a tile that transitions to \emph{Ready} or \emph{Inclusive} before fallback completion, or (2) another workgroup completing its fallback first, leaving the tile in \emph{Ready}. If the tile state is \emph{Inclusive} then the write must be skipped, especially because races in writes for the parts may leave the tile permanently in an unreadable state. Thus, a simple atomic store is not suitable. An atomic compare-and-swap would have the correct semantics, but WebGPU lacks this primitive (it does have \texttt{atomicCompareExchangeWeak}, which could possibly be made to work but makes reasoning about correctness and performance more difficult). Instead, we encode the tile state in the most significant bits of the atomic value and enforce a strict state ordering: \emph{Inclusive} $>$ \emph{Ready} $>$ \emph{Not Ready}. This allows an atomic \emph{maximum}, which successfully updates to a more advanced state but prevents a state regression, to correctly perform updates.

The atomic splitting technique generalizes to monoids of arbitrary size.
